---
title: "Control data clustering"
format: 
    html: default
date: "`r Sys.Date()`"
reference-location: margin
citation-location: margin
editor: visual
execute: 
  freeze: true
---

```{r}

library(Seurat)
library(stringr)
library(Rgb)
library(grid)
library(gridExtra)
library(viridis)
library(ggplot2)
library(stringr)
library(dendextend)
library(devtools)
source("../../IBD/code/raw_code/temp_functions.R")
load_all("../../../../COTAN/")
```

Really this should be the **COLITIS** (problem in the excel/geo consistency).

```{r}
seurat.data.control <- readRDS(file = "../data/03_clustered_data/control/Seurat_obj_control_with_cotan_clusters.RDS")
cotan.obj.control <- readRDS("../data/03_clustered_data/control.cotan.clustered.RDS")

dim(cotan.obj.control@cluster_data)
```

```{r}
cotan.obj.control@meta
```

# Seurat clustering

Using Seurat 17 clusters were identified.

```{r}
DimPlot(seurat.data.control, label = T, group.by = "seurat_clusters") + theme(legend.position="none")
```

# Cotan initial clusters

While, in the first round of clustering using COTAN the homogeneous clusters were 57 and 64 cells were dropped.

```{r}
DimPlot(seurat.data.control, label = T, group.by = "cotan") + theme(legend.position="none")
```

# Markers

This are the marker used also in the original paper.

| Cell type           | Markers                                                                                                                                                        |
|---------------------|----------------------------------------------------------------------------------------------------------------------------------------------------------------|
| B cells             | "MS4A1", "CD79A", "VPREB3", "CD37", "BANK1", "LY9"                                                                                                             |
| Plasma              | "JCHAIN", "IGHA2", "IGHA1", "IGKC"                                                                                                                             |
| T & NK cells        | "CCL5", "CD7", "CD3E", "CD3D", "NKG7", "CST7", "FYB", "CD8A", "HCST"                                                                                           |
| LTi                 | "IL22", "CSF2", "IL4I1", "XCL1", "KIT", "BHLHE40", "TNFRSF18", "KRT86", "NCOA7", "IL23R", "AQP3", "ALDOC", "TNFSF13B"                                          |
| Myeloid             | "LYZ", "CST3", "IL1B", "TPSAB1", "CXCL8", "CPA3", "TPSB2", "AIF1", "FCER1G", "C1QC", "C1QA", "C1QB", "S100A8"                                                  |
| Mitotic B & T cells | "STMN1","HMGB2", "HMGN2", "HIST1H4C", "H2AFZ","TUBB", "TUBA1B", "HMGB1", "UBE2C", "RRM2", "PTTG1", "DUT", "TYMS", "H2AFV", "MKI67", "TOP2A", "CENPF", "NUSAP1" |
| Fibro & Endo        | "CXCL14", "ADAMDEC1", "COL3A1", "APOE", "CFD", "IGFBP7", "CCL2", "COL1A2", "COL1A1", "A2M", "COL6A2", "IFITM3", "LUM", "DCN", "MFAP4"                          |
| Epithelial          | "PHGR1", "TFF3", "FABP1", "ELF3", "ZG16", "KRT8", "SPINK4", "LGALS4", "KRT18", "CLDN4", "PIGR", "S100A6", "FCGBP", "KRT19", "CLDN3"                            |

# Initial tree

```{r}
cluster_data <- cotan.obj.control@cluster_data

######## This is the best: cosine dissimilarity
Matrix <- as.matrix(t(cluster_data))
sim <- Matrix / sqrt(rowSums(Matrix * Matrix))
sim <- sim %*% t(sim)
D_sim <- as.dist(1 - sim)
tree <- hclust(D_sim,method = "ward.D2")

dend <- as.dendrogram(tree)
plot(dend)
```

# Final tree and markers enrichment

```{r}
cotan.obj.control <- readRDS("../data/03_clustered_data/control_merged_cotan.RDS")
cluster_data <- cotan.obj.control@cluster_data

######## This is the best: cosine dissimilarity
Matrix <- as.matrix(t(cluster_data))
sim <- Matrix / sqrt(rowSums(Matrix * Matrix))
sim <- sim %*% t(sim)
D_sim <- as.dist(1 - sim)
tree <- hclust(D_sim,method = "ward.D2")

dend <- as.dendrogram(tree)

```

```{r, fig.width=18}
#df <- read.csv("IBD_control_markers_clusters_percentage.csv",row.names = 1)
df <- read.csv("../data/04_cluster_identification/control_markers_clusters_expression.csv",row.names = 1)

dend <- as.dendrogram(tree)
df <- round(df[,dend %>% labels], digits = 2)
colnames(df) <- str_remove_all(colnames(df), pattern = "cl.")
cut = cutree(tree, k = 9)
dend =branches_color(dend,k=9,col=c("#4DBBD5FF","#91D1C2FF","#E64B35FF","gray80","#3C5488FF","#F39B7FFF","gray80"),groupLabels = T)
dend =color_labels(dend,k=9)
dd <- dend %>%
  dendextend::set("labels") %>%
  ggplot(horiz=F, axes=T)

#tab <- tableGrob(round(df[,dend %>% labels], digits = 0),theme = ttheme_minimal(base_size = 4))
tab <- grid.ftable(df, gp = gpar(fill = rep(c("grey80", "grey95"), each = (dim(df)[2]+1))),padding = unit(7.9, "mm"))

lay <- rbind(c(rep(1,10)),
               c(rep(2,9),NA))

#tab1 <- grid.ftable(matrix(rownames(df), c(1:length(rownames(df)))))

grid.arrange(dd, arrangeGrob(tab,widths = 1),  layout_matrix=lay)
```

# Final merged cluster in Seurat plot

```{r}
seurat.data.control <- readRDS(file = "../data/03_clustered_data/Seurat_obj_control_with_cotan_clusters_merged.RDS")
seurat.data.control <- SetIdent(seurat.data.control, value = 'cotan')
DimPlot(seurat.data.control, label = T) + theme(legend.position="none")
```

# Check for sample id

```{r}
seurat.data.control@meta.data$sample.id <- str_split(rownames(seurat.data.control@meta.data),pattern = "[.]",simplify = T)[,2]

DimPlot(seurat.data.control, label = T,group.by =  "sample.id", pt.size = 0.5)
```

The sample distribution is not perfect but it doesn't seems too bad...

# Cluster assignment and merge

```{r}
seurat.data.control@meta.data$Cell.type <- seurat.data.control@meta.data$cotan 

seurat.data.control@meta.data[seurat.data.control@meta.data$cotan %in% 
       c(36,18,17,34,40,6,24,38,1,31,47,12),]$Cell.type <- "Plama c."
seurat.data.control@meta.data[seurat.data.control@meta.data$cotan %in% 
       c(55),]$Cell.type <- "Mitotic Plasma c."
seurat.data.control@meta.data[seurat.data.control@meta.data$cotan %in% 
        c(19),]$Cell.type <- "Mitotic T c."
seurat.data.control@meta.data[seurat.data.control@meta.data$cotan %in% 
        c(3,46,2,43,33,29,4,7),]$Cell.type <- "Mitotic B c."
seurat.data.control@meta.data[seurat.data.control@meta.data$cotan %in% 
        c(14,0,28,30),]$Cell.type <- "B cells"
seurat.data.control@meta.data[seurat.data.control@meta.data$cotan %in% 
        c(5,11,10,25,9,32,8),]$Cell.type <- "T & NK cells"
seurat.data.control@meta.data[seurat.data.control@meta.data$cotan %in% 
        c(13),]$Cell.type <- "LTi"
seurat.data.control@meta.data[seurat.data.control@meta.data$cotan %in% 
        c(42,45,35,37),]$Cell.type <- "Myeloid"
seurat.data.control@meta.data[seurat.data.control@meta.data$cotan %in% 
        c(26,21,51,49,44),]$Cell.type <- "Fibro & Endo"
seurat.data.control@meta.data[seurat.data.control@meta.data$cotan %in% 
        c(53,54),]$Cell.type <- "Epitelial"
seurat.data.control <- SetIdent(seurat.data.control, value = 'Cell.type')

DimPlot(seurat.data.control, label = T,repel = T) + theme(legend.position="none")

```

```{r}
score.ISG <- read.csv("../data/04_cluster_identification/control_ISG_clusters_expression.csv",row.names = 1)

colnames(score.ISG) <- str_remove_all(colnames(score.ISG),pattern = "cl.")
seurat.data.control@meta.data$ISG.score <- NA

for (cl in unique(colnames(score.ISG))) {
  seurat.data.control@meta.data[seurat.data.control@meta.data$cotan == cl,]$ISG.score <- score.ISG[[cl]]
}  


FeaturePlot(seurat.data.control,features = "ISG.score",pt.size = 0.5,label.size = 4,repel = T,label = T,min.cutoff = 0, max.cutoff = 1)+
  scale_color_gradient2(low ="#EDFC1B", high = "#0B0074", 
                        mid = "#BC2E67",
                        midpoint = 0.5,limits = c(0,1))
```
